Response to reviewers

### 2025-09-03 paper_plos/paper_plos_V1.tex
final changes to R1

\section*{Abstract}
Transportation planning has long prioritized the efficiency of movement.
However, the concept of accessibility represents a more comprehensive
evolution, shifting focus from movement to the potential to reach--or
spatially interact with--desired destinations. Despite growing
recognition of accessibility-based planning approaches, the concept
remains fragmented, with inconsistent definitions and unclear
interpretations. To this end, this paper offers a methodological
contribution by specifying a family of accessibility measures grounded
in the shared theoretical roots of gravity-based accessibility and
spatial interaction models, particularly their balancing factors. From
this foundation, we outlined four memebers of the family: the
`unconstrained' measure (i.e., Hansen-type accessibility), the `total
constrained' measure (i.e., a constrained version of the Hansen-type
accessibility), the `singly constrained' measure (i.e., related to the
popular two step floating catchment approach - 2SFCA), and the `doubly
constrained' measure representing realized access (i.e., equal to the
doubly constrained spatial interaction model). These measures can be
interpreted as either the number of accessible opportunities or
accessible population (i.e., market potential). A toy example
illustrates how they produce interpretable, unit-based values, offering
a clearer and more coherent basis for accessibility analysis.

\linenumbers
\section{Introduction}\label{introduction}

@@

More specifically, the wide range of accessibility definitions, with
novel methods being more sophisticated but less intuitive
\citep{kapatsila_resolving_2023}, can further hinder practical uptake
\citep{vanweeAccessible2016}. Geurs and van Wee
\citep{geursAccessibilityEvaluationLanduse2004} classify accessibility
measures into four categories: infrastructure-, place-, person-, and
utility-based. Among place-based measures (this work's focus), variants
include gravity-based \citep{hansen1959}, cumulative opportunity
\citep{pirie_measuring_1979}, the 2 Step Floating Catchment Area (FCA)
approach \citep{luo2003}, and a variety of modifications to these
approaches e.g., Enhanced 2-Step FCA \citep{luoEnhanced2009}, 3-Stage
FCA \citep{wan2012three}, Modified 2-Step FCA
\citep{delamater2013spatial}, inverse 2-Step FCA
\citep{wang_2sfca_2021}, and n-steps FCA \citep{liang_novel_2024}. While
these methods are tailored to address specific research contexts,
overall this diversity does not demystify existing questions like those
raised in \citep{vanweeAccessible2016}: How should practitioners
interpret differences in accessibility scores between modes, and how
should results be communicated?

@@

In summary, this paper makes two contributions: (1) Reviews how spatial
interaction modeling and accessibility share similar origins, but
diverged in focus and interpretation. (2) Introduces a family of
accessibility measures grounded in spatial interaction principles,
including total, singly, and doubly constrained cases and variants of
``accessible opportunities'' and ``accessible population''. These cases
and variants align with common measures such as Hansen-type
accessibility \citep{hansen1959}, competitive accessibility measures
such as the 2SFCA method \citep{shen1998, luo2003}, and market potential
models
\citep{harris_market_1954, vickermanAccessibilityAttractionPotential1974}.

We contend that accessibility research should re-engage with spatial
interaction modeling, particularly the use of Wilson's system
constraints. While spatial interaction models embraced such constraints
to improve interpretability, most accessibility models have not. This
lack of adoption contributes to fuzziness in current analyses, limiting
interpretive clarity to simple proportional comparisons (e.g.,
``higher-than'', ``lower-than'')
\citep{millerAccessibilityMeasurementApplication2018}. Without such
constraints, accessibility scores lack clear units and comparability
across cities or modes. In contrast, constrained measures yield values
that can be tied to tangible values without any post-hoc treatment,
theoretically making them more interpretable, communicable, and
actionable in planning.

@@

This famous equation expresses that the attractive force \(F\) between
two bodies is directly proportional to the product of their masses and
inversely proportional to the square of the distance between them. As
mass increases, so does force; as distance increases, force decreases.
However, Eq~\ref{eq-phys-grav-prop} only demonstrates a
\emph{proportional} relationship. To quantify the magnitude (not the
\emph{proportional} magnitude) of the force, it must be
\emph{constrained} with an empirical constant. This constant \(G\)
converts Eq~\ref{eq-phys-grav-prop} from an expression of
proportionality to the following expression of equality:

@@

Hansen recast Stewart's population potential to reflect accessibility, a
model of human behaviour useful to capture regularities in mobility
patterns. Hansen replaced \(M_j\) in
Eq~\ref{eq-stewart-population-potential-sum} with \emph{opportunities}
to derive an \emph{opportunity potential}, or more specifically, a
\emph{potential of opportunities for interaction} as
\(S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}\) with a contemporary
rewriting of \(d^{-\beta}\) to \(f(d_{ij})\) accounts for a variety of
impedance functions beyond the inverse power as follows:

@@

However, Hansen's formulation carried forward a crucial omission that
continues to affect the literature: the proportionality constant \(G\)
included in Stewart's original formulation
(Eq~\ref{eq-stewart-population-potential-sum}) has vanished entirely.
Although Stewart included \(G\) explicitly (with a note that ``\(G\)
{[}was{]} left for future determination: a suitable choice of other
units can reduce it to unity'' {[}p.~34{]}.) Hansen made no mention of
it. As a result, modern accessibility analysis has largely evolved
without addressing the constant's role, leaving \(G\) effectively and
implicitly fixed at 1. This omission has significant implications.
Without a proportionality constant, the accessibility formula expresses
only a proportional relationship:
\(S_i \propto \sum_j g(O_j)f(d_{ij})\), not of calibrated equality.
Recognition of the nature of this relationship is not common in the
literature, but is known, i.e., this proportional equation is shown in
Figure 1 in Wu and Levinson \citep{wuUnifyingAccess2020}.

@@

While accessibility research evolved in North America with Hansen
\citep{hansen1959}, a parallel development was taking place across the
Atlantic with Alan G. Wilson. Wilson's ground breaking paper
\citep{wilson1971}, advanced a general framework for spatial interaction
modeling focused on flows of interaction between places and derived from
a non-Newtonian analogy. Additionally, this work was not focused on the
`potential' concept, as accessibility has been. Wilson
\citep{wilson1971} formalized the general spatial interaction model
through the following equation:

@@

Additional information can be introduced. For example, when information
is available about the total number of flows produced by each origin,
\(W_i^{(1)}\) in Eq~\ref{eq-phys-gravity-model} is represented as
\(O_i\) and the following constraint can be used:

@@

Depending on which of the three system constraints are applied a family
of spatial interaction models can be derived from
Eq~\ref{eq-phys-gravity-model}. The proportionality constant \(k\) is
replaced with different \emph{balancing factors}. This change in name is
more useful as these factors not only preserve proportionality but also
ensure that the predicted flows \(T_{ij}\) align with the known
constraints considered in the system. In other words, the different
balancing factors adjust the model so that flows become statistical
averages consistent with observed origin and/or destination data.

In the framework introduced and inferred from Wilson \citep{wilson1971},
three types of balancing factors are specified: (1) an unconstrained
model that only matches the total volume of interaction \(K\), (2) a
singly constrained model (either by origins \(A_i\) or destinations
\(B_j\)), and (3) a doubly constrained model that satisfies both.

@@

Wilson's work is notable for many reasons. Rather than relying on a
universal constant like \(G\) or a scaling factor to balance units,
Wilson's models calibrate interaction flows through known empirical
constraints using principles of entropy maximisation. This results in
interpretable, balanced (given the constraints), and unit-consistent
\(ij\) flows. In this way, spatial interaction modeling tradition can be
seen to have succeeded where accessibility modeling stalled. While
Wilson's model produces results that are in units of flow tethered to
the system of analysis, Hansen-type measures (still widely used today in
accessibility work) yield outputs that reflect proportional--not
equal--relationships and typically lack interpretable units.

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_files/figure-pdf/fig-docs-per-year-1.pdf}

@@

\noindent Where \(W_X^{(2)}\) and \(W_X^{(1)}\) is the mass of the
destination (i.e., opportunities) and mass of the origin (i.e.,
population) at either the zone or for the full region, depending on the
case (hence represented by a \(X\)).

@@

@@

Consider the simple 3-zone region in Figure
Fig~\ref{fig-analytical-device-conc-accessibility}, where each zone
serves as both origin (\(i\)) and destination (\(j\)). The system
includes three inputs: zonal population and opportunities, a zonal cost
(travel time) matrix, and three travel impedance functions representing
different travel behaviours.

First, Table Table~\ref{tbl-small-system-land-use} summarizes population
(in 10,000s) and physicians per zone. For context, the
Provider-to-Population Ratio (PPR) is 24.5 comparable to Canada's 2022
PPR of 24.97 physicians per 10,000 \citep{whoMedicalDoctors102025}.
Second, Table Table~\ref{tbl-small-system-cost} shows travel times
(minutes); it can be discerned that Zones 1 and 3 are closer to each
other than to Zone 2. Zones 1 and 3 together have a population roughly
equal to Zone 2 but offer more than twice the physician availability. We
interpret Zone 1 as the Urban Edge, Zone 3 as part of the Urban Core,
and Zone 2 as the Suburban.

@@

For example, Table
Table~\ref{tbl-simple-example-unconstrained-accessibility} shows
\(V^0_{i}\) under each decay function. Comparing across decay types is
meaningless in absolute terms. For instance, the difference in zone 1
(edge of urban core)'s accessibility under \(f_3\) vs \(f_1\) is 370.92,
but in what units? These two values are a product of different impedance
functions ( \emph{physicians-minute\(^{-3}\)} and
\emph{physicians-minute\(^{-0.1}\)}), making the comparison
uninterpretable (and arguably incorrect). The fundamental
uninterpretability of what is a
\emph{opportunity-weighted-travel-impedance} unit remains.

@@

\begin{itemize}
\tightlist
\item
  \(V_i^T\): accessibility is constrained by the total number of
  opportunities (total constrained accessible opportunity) and which is
  interpreted as Hansen's accessibility with a constraining constant,
  and
\item
  \(M_j^T\): where \(i\) and \(j\) of the first variant is transposed,
  yielding a measure constrained by the total number of population and
  to be interpreted as constrained `market potential'.
\end{itemize}

@@

In the total constrained case, accessibility is expressed as a share of
the total number of opportunities in the region \(D\), allocated based
on travel impedance. The total constrained accessibility from \(i\) to
\(j\):
\begin{equation}\phantomsection\label{eq-total-constrained-access}{
V^T_{ij} = \kappa_{ij}^T D
}\end{equation}

@@

Referring back to our simple numeric example, \(K^T\) for the highest
decay t scenario \(f_1(c_{ij}) = 1/c_{ij}^3\) would then be:

@@

\(K^T\) for other decay scenarios are calculated similarly in code.
Applying each \(K^T\) to the unconstrained values \(V_i^0\) yields total
constrained accessibility values (Table
Table~\ref{tbl-simple-example-total-opportunity-accessibility}), all in
units of physicians.

@@

Compared to the unconstrained case, values now sum to the known regional
total \(D\), allowing interpretation of absolute and relative
differences across zones and travel scenarios. For example, in the
highest decay case, Zone 1 (Urban Edge) captures an intermediate number
of physicians (172.0652825), like in the unconstrained accessibility
case. However, unlike in the unconstrained case, we can say that this
value is out of the 490 physicians in the region, which allows us also
to deduce that zone 1 captures 1.3072213 and 0.9235529 times more than
zone 2 and 3. Values for the lesser decay (\(f_2(c_{ij})\)) and lowest
decay (\(f_3(c_{ij})\)) scenarios are calculated separately, with decay
scenario values also summing to equal 490 physicians accessible in the
region.

One can also directly compare values at a specific zone, across travel
impedance scenarios, due to the consistent units. As the decay scenario
decreases, all zones become more accessible to each other and the
differences between pairs diminishes (i.e., in \(f_3(c_{ij})\) each zone
captures close to an average amount of physicians, a third of 490 or
\textasciitilde163). In terms of proportional magnitude, this can also
be observed in the unconstrained measure for this scenario. However, for
the total constrained measure, this plateauing of results have meaning.
In fact, each zone is allocated an average of the total amount in the
region, as a result of the total constrained proportional allocation
factor.

However, what's notable is how Zones change between scenarios. For
instance, Zone 1's share only declines slightly--these declines are
outpaced by Zone 2's relative gains. This shift reflects how
\(\kappa_{ij}^T\) redistributes opportunities in proportion to impedance
under different travel behaviours.

@@

This second total constrained variant is a transpose of the
opportunity-constrained formulation, switching indices \(i\) and \(j\)
to yield s measure of market potential--the number of people who can
spatially interact with a destination.

@@

This formulation has been referred to as \textbf{spatial availability}
by Soukhov et al. \citep{soukhovIntroducingSpatialAvailability2023},
since it incorporates spatial competition by allocating opportunities
based on demand (population), impedance, and the known opportunity
totals \(D_j\). The dimensionless factor \(\kappa^S_{ij}\) ensures that
each destination's opportunities are distributed proportionally to
origins. As in the total constrained case, \(V_i^S\) is expressed in the
units of accessible opportunities. Finally, the \textbf{per capita}
version of this measure:

Soukhov et al., 2023 \citep{soukhovIntroducingSpatialAvailability2023}
also showed that the following expression (accessibility per capita) is
a constrained version of the popular 2SFCA approach of Shen 1998
\citep{shen1998} and Luo and Wang 2003 \citep{luo2003}:

@@

The balancing factors \(B_j\) for the \(f_2(c_{ij})\) decay group for
zones 1, 2 and 3 area 12.8571429, 8.7685113 and 10.6635071,
respectively. For the \(f_3(c_{ij})\) decay group, they are 0.0672461,
0.0660559 and 0.0663798. Using these these balancing constants, we can
calculate the singly constrained opportunity accessibility
(Table~\ref{tbl-simple-example-attraction-constrained-accessibility}).

@@

This consideration has important implications. Consider the highest
decay scenario \(f_1(c_{ij})\). Under this scenario, Zone 1--despite
hosting a medium amount of physicans--captures the fewest physicians
(133.4687282), compared to 166.7813387 at Zone 2, and 189.7499331 at
Zone 3. Why? Zone 1 has the smallest population and is adjacent to Zone
3, the urban core. Its low impedance-weighted demand means
\(\kappa^S_{ij}\) allocates it fewer opportunities. By contrast, in the
total constrained case, Zone 1 fares better, capturing 35\% of all
physicians (compared to 27\%).

@@

In this way, the consideration of constrained accessibility \emph{per
capita} may be clarifying. Often, accessibility values are reported as
raw scores without the consideration for population. But, as we
introduced constraints, these constrained accessibility values can be
normalized using anything that is relevant to the zone. In
Table~\ref{tbl-simple-example-attraction-constrained-accessibility-per-capita},
we present per capita accessibility for the numeric example, simply in
units of number of physicians accessible per population at each zone.
Notably, these per capita rates are equivalent to the 2SFCA values.

@@

Similar to Eq~\ref{eq-total-population-balancing-factor} in transposing
the origins and destinations, we can define a \emph{singly constrained}
measure of market potential that preserves the known population (i.e.,
the mass weight at the origin \(W_i^{(1)}\) is now represented by
\(O_i\)). In it's per-capita expression, i.e., equivalent to 2SFCA, this
constrained concept of market potential been used to express ``facility
crowdedness'' as in Wang \citep{wang_inverted_2018}.

@@

As the doubly constrained accessibility measure \(V_{ij}^D\) takes the
form of the production-attraction (doubly constrained) spatial
interaction model as follows:

@@

Moving onto the toy example, to calculate doubly constrained
accessibility, the interpretation of the population data and the counts
of the opportunity data in the numeric example must be reinterpreted.
Namely, a count of physician \emph{capacity} per destination \(D_j\) is
needed instead of simply the number of physicians. Also, we must be able
to clearly state that the population is the \emph{capacity} of the
origin to interact with opportunities \(O_i\), i.e., the count of people
seeking opportunities.

This adjusted simple example is summarised in
Table~\ref{tbl-small-system-land-use-doubly-constrained-case}: with the
population (in units of 10,000s of people seeking physicians) and the
opportunities (in units of 10,00s of physician-capacity) per zone. For
the population, we leave this unchanged numerically but we now must keep
in mind that each person interacts with one physician capacity. The
number of providers per destination is however revised to represent
physician capacity, scaled approximately from the original number of
physicians used in previous cases
(Table~\ref{tbl-small-system-land-use}). The system-wide PPR is now:1,
this is compared to the unmodified example which yields system PPR
of24.5.

@@

As shown in
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}
for the highest-decay scenario \(f_1(c_{ij})\), accessibility is no
longer meaningfully represented as zonal summaries like \(V^D_i\) or
\(M^D_j\), since these values reproduce the original constraints--i.e.,
\(V^D_i = O_i\) so the physician-capacity accessible for Zones 1, 2, and
3 would be 4.0018381, 9.9963656, and 6.0017963. Hence, the usefulness of
the doubly constrained measure lies in the interpretation as
\(V_{ij}^D\) values. \(V^D_{ij}\) values represent the number of
opportunities from zone \(j\) allocated to populations in zone \(i\),
shaped by both mass and travel impedance.

To illustrate this, consider the results for Zone 2 (Suburban zone--a
high-population, low-opportunity, relatively remote zone). As shown in
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2},
its intrazonal allocation (i.e., \(V^D_{22}\)) declines as travel
impedance decay decreases--from 4.9593248 under \(f_1(c_{ij})\) to
2.6672837 under \(f_3(c_{ij})\), out of the \textasciitilde10
opportunities allocated to Zone 2 (a population of 10).

>>>>

\section*{Abstract}
Transportation planning has long prioritized the efficiency of movement.
However, the concept of accessibility represents a more comprehensive
evolution, shifting focus from movement to the potential to reach--or
spatially interact with--desired destinations. Despite growing
recognition of accessibility-based planning approaches, the concept
remains fragmented, with inconsistent definitions and unclear
interpretations. To this end, this paper offers a methodological
contribution by specifying a family of accessibility measures grounded
in the shared theoretical roots of gravity-based accessibility and
spatial interaction models, particularly their balancing factors. From
this foundation, we outline four memebers of the family: the
`unconstrained' measure (i.e., Hansen-type accessibility), the `total
constrained' measure (i.e., a constrained version of the Hansen-type
accessibility), the `singly constrained' measure (i.e., related to the
popular two step floating catchment approach - 2SFCA), and the `doubly
constrained' measure representing realized access (i.e., equal to the
doubly constrained spatial interaction model). These measures can be
interpreted as either the number of accessible opportunities or
accessible population (i.e., market potential). A toy example
illustrates how they produce interpretable, unit-based values, offering
a clearer and more coherent basis for accessibility analysis.

\linenumbers

\section{Introduction}\label{introduction}

@@

More specifically, the wide range of accessibility definitions, with
novel methods being more sophisticated but less intuitive
\citep{kapatsila_resolving_2023}, can further hinder practical uptake
\citep{vanweeAccessible2016}. Geurs and van Wee
\citep{geursAccessibilityEvaluationLanduse2004} classify accessibility
measures into four categories: infrastructure-, place-, person-, and
utility-based. Among place-based measures (this work's focus), variants
include gravity-based \citep{hansen1959}, cumulative opportunity
\citep{pirie_measuring_1979}, the 2 Step Floating Catchment Area (FCA)
approach \citep{luo2003}, and a variety of modifications to these
approaches e.g., Enhanced 2-Step FCA \citep{luoEnhanced2009}, 3-Stage
FCA \citep{wan2012three}, Modified 2-Step FCA
\citep{delamater2013spatial}, inverse 2-Step FCA
\citep{wang_2sfca_2021}, and n-steps FCA \citep{liang_novel_2024}. While
these methods are tailored to address specific research contexts,
overall this diversity does not demystify existing questions like those
raised by van Wee \citep{vanweeAccessible2016}: How should practitioners
interpret differences in accessibility scores between modes, and how
should results be communicated?

@@

In summary, this paper makes two contributions: (1) we review how
spatial interaction modeling and accessibility share similar origins but
have diverged in focus and interpretation. (2) we introduce a family of
accessibility measures grounded in spatial interaction principles,
including total, singly, and doubly constrained cases and variants of
``accessible opportunities'' and ``accessible population''. These cases
and variants align with common measures such as Hansen-type
accessibility \citep{hansen1959}, competitive accessibility measures
such as the 2SFCA method \citep{shen1998, luo2003}, and market potential
models
\citep{harris_market_1954, vickermanAccessibilityAttractionPotential1974}.

We contend that accessibility research should re-engage with spatial
interaction modeling, particularly the use of Wilson's
\citep{wilson1971} system constraints. While spatial interaction models
embraced such constraints to improve interpretability, most
accessibility models have not. This lack of adoption contributes to
fuzziness in current analyses, limiting interpretive clarity to simple
proportional comparisons (e.g., ``higher-than'', ``lower-than'')
\citep{millerAccessibilityMeasurementApplication2018}. Without such
constraints, accessibility scores lack clear units and comparability
across cities or modes. In contrast, constrained measures yield values
that can be tied to tangible values without any post-hoc treatment,
theoretically making them more interpretable, communicable, and
actionable in planning.

@@

This famous equation expresses that the attractive force \(F\) between
two bodies \(i\) and \(j\) is directly proportional to the product of
their masses and inversely proportional to the square of the distance
between them. As mass increases, so does force; as distance increases,
force decreases. However, Eq~\ref{eq-phys-grav-prop} only demonstrates a
\emph{proportional} relationship. To quantify the magnitude (not the
\emph{proportional} magnitude) of the force, it must be
\emph{constrained} with an empirical constant. This constant \(G\)
converts Eq~\ref{eq-phys-grav-prop} from an expression of
proportionality to the following expression of equality:

@@

Hansen recast Stewart's population potential to reflect accessibility, a
model of human behaviour useful to capture regularities in mobility
patterns. Hansen replaced \(M_j\) in
Eq~\ref{eq-stewart-population-potential-sum} with \emph{opportunities}
to derive an \emph{opportunity potential}, or more specifically, a
\emph{potential of opportunities for interaction} as
\(S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}\). A contemporary rewriting
of \(d^{-\beta}\) to \(f(d_{ij})\) accounts for the variety of impedance
functions beyond the inverse power used in the applied literature:

@@

However, Hansen's formulation carried forward a crucial omission that
continues to affect the literature: the proportionality constant \(G\)
included in Stewart's original formulation
(Eq~\ref{eq-stewart-population-potential-sum}) has vanished entirely.
Although Stewart included \(G\) explicitly (with a note that ``\(G\)
{[}was{]} left for future determination: a suitable choice of other
units can reduce it to unity'' {[}p.~34{]}.) Hansen made no mention of
it. As a result, modern accessibility analysis has largely evolved
without addressing the constant's role, leaving \(G\) effectively and
implicitly fixed at 1. This omission has significant implications.
Without a proportionality constant, the accessibility formula expresses
only a proportional relationship:
\(S_i \propto \sum_j g(O_j)f(d_{ij})\), not one of calibrated equality.
Recognition of the nature of this relationship is not common in the
literature, but is known, i.e., this proportional equation is shown in
Figure 1 in Wu and Levinson \citep{wuUnifyingAccess2020}.

@@

While accessibility research evolved in North America with Hansen
\citep{hansen1959}, a parallel development was taking place across the
Atlantic with Alan G. Wilson. Wilson's ground breaking paper
\citep{wilson1971} advanced a general framework for spatial interaction
modeling focused on flows of interaction between places and derived from
a non-Newtonian analogy. Additionally, this work was not focused on the
`potential' concept, as accessibility has been. Wilson
\citep{wilson1971} formalized the general spatial interaction model
through the following equation:

@@

Additional information can be introduced. For example, when information
is available about the total number of flows produced by each origin,
\(W_i^{(1)}\) in Eq~\ref{eq-phys-gravity-model}, represented as \(O_i\),
then following constraint can be used:

@@

Depending on which of the three system constraints are applied, a family
of spatial interaction models can be derived from
Eq~\ref{eq-phys-gravity-model}. The proportionality constant \(k\) is
replaced with different \emph{balancing factors}. This change in name is
more useful as these factors not only preserve proportionality but also
ensure that the predicted flows \(T_{ij}\) align with the known
constraints considered in the system. In other words, the different
balancing factors adjust the model so that flows become statistical
averages consistent with observed origin and/or destination data.

In the framework introduced and inferred from Wilson \citep{wilson1971},
three types of balancing factors are specified: (1) an unconstrained
model that only matches the total volume of interaction \(K\), (2) a
singly constrained model (either by origins \(A_i\) or destinations
\(B_j\) - see below), and (3) a doubly constrained model that satisfies
both.

@@

Wilson's work is notable for many reasons. Rather than relying on a
universal constant like \(G\) or a scaling factor to balance units,
Wilson's models calibrate interaction flows through known empirical
constraints using principles of entropy maximisation. This results in
interpretable, balanced (given the constraints), and unit-consistent
\(ij\) flows. In this way, spatial interaction modeling tradition can be
seen to have succeeded where accessibility modeling stalled. While
Wilson's model produces results that are in units of flow tethered to
the system of analysis, Hansen-type measures (still widely used today in
accessibility work) yield outputs that reflect proportional -- not equal
-- relationships and typically lack interpretable units.

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{paper_plos_V1_files/figure-pdf/fig-docs-per-year-1.pdf}

@@

\noindent Where \(W_X^{(2)}\) and \(W_X^{(1)}\) reflect the mass of the
destination (i.e., opportunities) and mass of the origin (i.e.,
population) at either the zone or for the full region, depending on the
case (hence represented by a \(X\)).

@@

As a summary, each member of the family of accessibility measure is
named, explained in plain language, alongside their balancing factor(s),
proportional allocation factor(s), and mathematical equation and value
interpretations in Table 1.

\blandscape

Table 1: Summary of constrained accessibility measure types and
interpretations

\input{summary-family-access-measures-table.tex}

\elandscape

@@

Consider the simple 3-zone region in
Fig~\ref{fig-analytical-device-conc-accessibility}, where each zone
serves as both origin (\(i\)) and destination (\(j\)). The system
includes three inputs: zonal population and opportunities, a zonal cost
(travel time) matrix, and three travel impedance functions representing
different travel behaviours.

First, Table~\ref{tbl-small-system-land-use} summarizes population (in
10,000s) and physicians per zone. For context, the
Provider-to-Population Ratio (PPR) is 24.5 comparable to Canada's 2022
PPR of 24.97 physicians per 10,000 \citep{whoMedicalDoctors102025}.
Second, Table~\ref{tbl-small-system-cost} shows travel times (minutes);
it can be discerned that Zones 1 and 3 are closer to each other than to
Zone 2. Zones 1 and 3 together have a population roughly equal to Zone 2
but offer more than twice the physician availability. We interpret Zone
1 as the Urban Edge, Zone 3 as part of the Urban Core, and Zone 2 as
Suburban.

@@

For example, Table~\ref{tbl-simple-example-unconstrained-accessibility}
shows \(V^0_{i}\) under each decay function. Comparing across decay
types is meaningless in absolute terms. For instance, the difference in
zone 1 (edge of urban core)'s accessibility under \(f_3\) vs \(f_1\) is
370.92, but in what units? These two values are a product of different
impedance functions ( \emph{physicians-minute\(^{-3}\)} and
\emph{physicians-minute\(^{-0.1}\)}), making the comparison
uninterpretable (and arguably incorrect). The fundamental
uninterpretability of what is an
\emph{opportunity-weighted-travel-impedance} unit remains.

@@

\begin{itemize}
\tightlist
\item
  \(V_i^T\): accessibility is constrained by the total number of
  opportunities (total constrained accessible opportunity) and which is
  interpreted as Hansen's accessibility with a constraining constant,
  and
\item
  \(M_j^T\): where \(i\) and \(j\) of the first variant is transposed,
  yielding a measure constrained by the total population and to be
  interpreted as constrained `market potential'.
\end{itemize}

@@

In the total constrained case, accessibility is expressed as a share of
the total number of opportunities in the region \(D\), allocated based
on travel impedance. The total constrained accessibility from \(i\) to
\(j\) takes the form:
\begin{equation}\phantomsection\label{eq-total-constrained-access}{
V^T_{ij} = \kappa_{ij}^T D
}\end{equation}

@@

Referring back to our simple numeric example, \(K^T\) for the highest
decay travel scenario \(f_1(c_{ij}) = 1/c_{ij}^3\) would then be:

@@

\(K^T\) for other decay scenarios are calculated similarly in code.
Applying each \(K^T\) to the unconstrained values \(V_i^0\) yields total
constrained accessibility values
(Table~\ref{tbl-simple-example-total-opportunity-accessibility}), all in
units of physicians.

@@

Compared to the unconstrained case, values now sum to the known regional
total \(D\), allowing interpretation of absolute and relative
differences across zones and travel scenarios. For example, in the
highest decay case, Zone 1 (Urban Edge) captures an intermediate number
of physicians (172.065), like in the unconstrained accessibility case.
However, unlike in the unconstrained case, we can say that this value is
out of the 490 physicians in the region, which allows us also to deduce
that zone 1 captures 1.307 and 0.924 times more than zone 2 and 3.
Values for the lesser decay (\(f_2(c_{ij})\)) and lowest decay
(\(f_3(c_{ij})\)) scenarios are calculated separately, with decay
scenario values also summing to equal 490 physicians accessible in the
region.

One can also directly compare values at a specific zone, across travel
impedance scenarios, due to the consistent units. As the decay scenario
decreases, all zones become more accessible to each other and the
differences between pairs diminishes (i.e., in \(f_3(c_{ij})\) each zone
captures close to an average amount of physicians, a third of 490 or
\textasciitilde163). In terms of proportional magnitude, this can also
be observed in the unconstrained measure for this scenario. However, for
the total constrained measure, this plateauing of results has meaning.
In fact, each zone is allocated an average of the total amount in the
region, as a result of the total constrained proportional allocation
factor.

However, what's notable is how zones change between scenarios. For
instance, Zone 1's share only declines slightly--these declines are
outpaced by Zone 2's relative gains. This shift reflects how
\(\kappa_{ij}^T\) redistributes opportunities in proportion to impedance
under different travel behaviours.

@@

This second total constrained variant is a transpose of the
opportunity-constrained formulation, switching indices \(i\) and \(j\)
to yield a measure of market potential--the number of people who can
spatially interact with a destination.

@@

This formulation has been referred to as \textbf{spatial availability}
by Soukhov et al. \citep{soukhovIntroducingSpatialAvailability2023},
since it incorporates spatial competition by allocating opportunities
based on demand (population), impedance, and the known opportunity
totals \(D_j\). The dimensionless factor \(\kappa^S_{ij}\) ensures that
each destination's opportunities are distributed proportionally to
origins. As in the total constrained case, \(V_i^S\) is expressed in the
units of accessible opportunities.

Soukhov et al. \citep{soukhovIntroducingSpatialAvailability2023} also
showed that the following expression (accessibility per capita) is a
constrained version of the popular 2SFCA approach of Shen
\citep{shen1998} and Luo and Wang \citep{luo2003}:

@@

The balancing factors \(B_j\) for the \(f_2(c_{ij})\) decay group for
zones 1, 2 and 3 are 12.857, 8.769 and 10.664, respectively. For the
\(f_3(c_{ij})\) decay group, they are 0.067, 0.066 and 0.066. Using
these these balancing constants, we can calculate the singly constrained
opportunity accessibility
(Table~\ref{tbl-simple-example-attraction-constrained-accessibility}).

@@

This consideration has important implications. Consider the highest
decay scenario \(f_1(c_{ij})\). Under this scenario, Zone 1--despite
hosting a medium amount of physicians--captures the fewest physicians
(133.469), compared to 166.781 at Zone 2, and 189.75 at Zone 3. Why?
Zone 1 has the smallest population and is adjacent to Zone 3, the urban
core. Its low impedance-weighted demand means \(\kappa^S_{ij}\)
allocates it fewer opportunities. By contrast, in the total constrained
case, Zone 1 fares better, capturing 35\% of all physicians (compared to
27\%).

@@

In this way, the consideration of constrained accessibility \emph{per
capita} may be clarifying. Often, accessibility values are reported as
raw scores without considering the population potentially accessing
them. But, as we introduced constraints, these constrained accessibility
values can be normalized using anything that is relevant to the zone. In
Table~\ref{tbl-simple-example-attraction-constrained-accessibility-per-capita},
we present per capita accessibility for the numeric example simply in
units of number of physicians accessible per population at each zone.
Notably, these per capita rates are equivalent to the 2SFCA values.

@@

Similar to Eq~\ref{eq-total-population-balancing-factor} in transposing
the origins and destinations, we can define a \emph{singly constrained}
measure of market potential that preserves the known population (i.e.,
the mass weight at the origin \(W_i^{(1)}\) is now represented by
\(O_i\)). In it's per-capita expression, i.e., equivalent to 2SFCA, this
constrained concept of market potential has been used to express
``facility crowdedness'' as in Wang \citep{wang_inverted_2018}.

@@

The doubly constrained accessibility measure \(V_{ij}^D\) takes the form
of the production-attraction (doubly constrained) spatial interaction
model as follows:

@@

Moving on to the toy example, to calculate doubly constrained
accessibility, the interpretation of the population data and the counts
of the opportunity data in the numeric example must be reinterpreted.
Namely, a count of physician \emph{capacity} per destination \(D_j\) is
needed instead of simply the number of physicians. Also, we must be able
to clearly state that the population is the \emph{capacity} of the
origin to interact with opportunities \(O_i\), i.e., the count of people
seeking opportunities.

This adjusted simple example is summarised in
Table~\ref{tbl-small-system-land-use-doubly-constrained-case}. With the
population (in units of 10,000s of people seeking physicians) and the
opportunities (in units of 10,000s of physician-capacity) per zone. For
the population, we leave this unchanged numerically but we now must keep
in mind that each person interacts with one physician capacity. The
number of providers per destination is however revised to represent
physician capacity, scaled approximately from the original number of
physicians used in previous cases
(Table~\ref{tbl-small-system-land-use}). The system-wide PPR is now 1,
this is compared to the unmodified example which yields system PPR of
24.5.

@@

As shown in
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}
for the highest-decay scenario \(f_1(c_{ij})\), accessibility is no
longer meaningfully represented as zonal summaries like \(V^D_i\) or
\(M^D_j\), since these values reproduce the original constraints--i.e.,
\(V^D_i = O_i\) so the physician-capacity accessible for Zones 1, 2, and
3 would be 4.002, 9.996, and 6.002. Hence, the usefulness of the doubly
constrained measure lies in the interpretation as \(V_{ij}^D\) values.
\(V^D_{ij}\) values represent the number of opportunities from zone
\(j\) allocated to populations in zone \(i\), shaped by both mass and
travel impedance.

To illustrate this, consider the results for Zone 2 (Suburban zone--a
high-population, low-opportunity, relatively remote zone). As shown in
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2},
its intrazonal allocation (i.e., \(V^D_{22}\)) declines as travel
impedance decay decreases--from 4.959 under \(f_1(c_{ij})\) to 2.667
under \(f_3(c_{ij})\), out of the \textasciitilde10 opportunities
allocated to Zone 2 (a population of 10).
